---
title: "RStudio Intro"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: "Introduction to R and RStudio"
---

```{r setup}
#| include: false

library(learnr)
library(ai4teaching) # isntall with: devtools::install_github('johnsonra/ai4teaching')
Sys.setenv(model = 'gemini-1.5-pro')
```

The exercises in this document are a companion to [this tutorial](https://youtu.be/_98f-UMBMz4?si=HhszKrM9yNmyQQ04), which follows along closely with [Lesson 1](https://swcarpentry.github.io/r-novice-gapminder/01-rstudio-intro.html) of [R for Reproducible Scientific Analysis](https://swcarpentry.github.io/r-novice-gapminder/), provided by the [Software Carpentry](https://software-carpentry.org/) community under the creative commons license, [CC-BY 4.0](https://creativecommons.org/licenses/by/4.0/). For more information see [The Carpentries](https://carpentries.org/).

::: callout-tip
## Key Points

-   Use RStudio to write and run R programs.
-   R has the usual arithmetic operators and mathematical functions.
-   Use `<-` to assign values to variables.
-   Use `ls()` to list the variables in a program.
-   Use `rm()` to delete objects in a script.
-   Use `install.packages()` to install packages and `library()` to load them.
:::


#### Challenge 1

```{r challenge1, echo=FALSE}
question("Which of the following are valid R variable names?",
         answer('min_height', correct = TRUE),
         answer('max.height', correct = TRUE),
         answer('_age'),
         answer('.mass', correct = TRUE),
         answer('MaxLength', correct = TRUE),
         answer('min-length'),
         answer('2widths'),
         answer('celsius2kelvin', correct = TRUE),
         allow_retry = TRUE)
```


#### Challenge 2

```{r}
#| echo: true
mass <- 47.5
age <- 122
mass <- mass * 2.3
age <- age - 20
```

```{r challenge2, echo=FALSE}
quiz(question("What is the value of `mass` after running the lines of code above?",
              answer(109.25, correct = TRUE),
              type = "learnr_text"),
     question("What is the value of `age` after running the lines of code above?",
              answer(102, correct = TRUE),
              type = "learnr_text"),
     caption = '')
```


#### Challenge 3

```{r challenge3, echo=FALSE}
quiz(question_text("Run the code from the previous challenge, and write a command to check if `mass` is larger than `age`.",
                   answer_fn(function(value){
                     check_answer("Given two variables, `mass` and `age`, provide a line of R code that will check if `mass` is greater than `age`",
                                       value,
                                       model = Sys.getenv('model'),
                                       api_key = OPsecrets::get_secret('GEMINI_API_KEY', 'Private', 'Gemini', 'api_key'))
                   }), allow_retry = TRUE),
     question("Is mass larger than age?",
              answer('Yes', correct = TRUE),
              answer('No')),
     caption = '')
```


#### Challenge 4

```{r challenge4, echo=FALSE}
question_text("Write a command to remove the `mass` and `age` variables from your workspace.",
              answer_fn(function(value){
                check_answer("Provide a line of R code that will remove the variables `mass` and `age` from the workspace.",
                             value,
                             model = Sys.getenv('model'),
                             api_key = OPsecrets::get_secret('GEMINI_API_KEY', 'Private', 'Gemini', 'api_key'))
                }), allow_retry = TRUE)
```


#### Challenge 5

```{r challenge5, echo=FALSE}
question_text("Provide a command that will install the `gapminder` package.",
              answer_fn(function(value){
                check_answer("Provide the R command that will install the `gapminder` package.",
                             value,
                             model = Sys.getenv('model'),
                             api_key = OPsecrets::get_secret('GEMINI_API_KEY', 'Private', 'Gemini', 'api_key'))
                }), allow_retry = TRUE)
```
