---
title: "Vectorization"
format: html
editor: visual
---

The exercises in this document follow along closely with [Lesson 9](https://swcarpentry.github.io/r-novice-gapminder/09-vectorization.html) of [R for Reproducible Scientific Analysis](https://swcarpentry.github.io/r-novice-gapminder/), provided by the [Software Carpentry](https://software-carpentry.org/) community under the creative commons license, [CC-BY 4.0](https://creativecommons.org/licenses/by/4.0/). For more information see [The Carpentries](https://carpentries.org/).

::: callout-tip
## Key Points

-   Use vectroized options instead of loops
:::

```{r}
library(gapminder)
data(gapminder)

library(ggplot2)
library(cowplot)
theme_set(theme_cowplot())

library(magrittr)
```

#### Challenge 1

Make a new column in the `gapminder` data frame that contains population in units of millions of people. Check the head or tail of the data frame to make sure it worked.

<!-- Challenge 1 answer begin -->

```{r}
gapminder$pop_mill <- gapminder$pop / 1e6
```

<!-- Challenge 1 answer end -->

#### Challenge 2a

On a single graph, plot population, in millions, against year, for all countries. Do not worry about identifying which country is which.

<!-- Challenge 2a answer begin -->

```{r}
ggplot(gapminder, aes(year, pop_mill)) +
  geom_point() +
  labs(x = 'Year',
       y = 'Population (millions)')
```

<!-- Challenge 2a answer end -->

#### Challenge 2b

Repeat the exercise, graphing only for China, India, and Indonesia. Again, do not worry about which is which.

<!-- Challenge 2b answer begin -->

```{r}
gapminder[gapminder$country %in% c('China', 'India', 'Indonesia'),] %>%
  ggplot(aes(year, pop_mill)) +
  geom_point() +
  labs(x = 'Year',
       y = 'Population (millions)')
```

<!-- Challenge 2b answer end -->

#### Challenge 3

Given the following matrix:

```{r}
m <- matrix(1:12, nrow=3, ncol=4)
m
```

Write down what you think will happen when you run:

-   `m ^ -1`
-   `m * c(1, 0, -1)`
-   `m > c(0, 20)`
-   `m * c(1, 0, -1, 2)`

Did you get the output you expected? If not and the video didn't clear things up, let me know and we can discuss.

<!-- Challenge 3 answer begin -->

-   `m^-1` divides all the values in `m` by 1.

-   `m * c(1, 0, -1)` will recycle the values in the vector on the right, resulting in row 1 being multiplied by 1, row two being multiplied by 0, and row 3 being multiplied by -1.

-   `m > c(0, 20)` will recycle the values in the vector on the right. The values in `m` will be compared one column at a time. Since it is of length 2, this will not result in the same comparison being made across each row.

-   `m * c(1, 0, -1, 2)` will recycle the values in the vector on the right. The values in `m` will be compared one column at a time, so the first column will be multiplied by `c(1, 0, -1)` and the first value of the second column will be multiplied by 2. The next two values of column 2 will be multiplied by `c(1, 0)` with the first two values of column 3 being multiplied by `c(-1, 2)`, and so forth.

<!-- Challenge 3 answer end -->

#### Challenge 4a

We're interested in looking at the sum of the following sequence of fractions:

```         
x = 1/(1^2) + 1/(2^2) + 1/(3^2) + ... + 1/(n^2)
```

This would be tedious to type out, and impossible for high values of $n$. Use vectorization to compute $x$ when $n=100$.

<!-- Challenge 4a answer begin -->

```{r}
sum( 1 / (1:100)^2 )
```

<!-- Challenge 4a answer end -->

#### Challenge 4b

What is the sum when $n=10,000$?

<!-- Challenge 4b answer begin -->

```{r}
n <- 1e4

sum(1 / (1:n)^2 )
```

<!-- Challenge 4b answer end -->
